---
title: "**Buzzing Threats: Climate Change, Mosquitoes, and West Nile Virus in New York City**"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
library(httr)
library(GSODR)
library(sf)
library(patchwork)
```

## **Motivation**
merging motivaion (project goals, motivation) and related work (things that inspired us)
  
  
## **Related Work**

Our group has a shared interest in addressing climate change, studying infectious diseases, and uniting over our mutual hatred of mosquito bites. Inspiration for this project stems from several relevant news articles and papers:

* A [2024 New York Times article](https://www.nytimes.com/article/west-nile-virus-nyc.html) highlights the rising concern for an increased number of mosquitoes infected with West Nile virus (WNV) over the summer in New York City (NYC) and how this likely driven by climate change. 
* A [2020 Nature article](https://www.nature.com/articles/s41590-020-0648-y) finds that climate change can affect mosquito abundance, survival, transmission dynamics, and pathogen development within vectors.

This issue is concerning as worsening climate change trends and warmer temperatures in NYC can create more hospitable environments for mosquitoes carrying WNV and may increase human-mosquito interaction. Motivated by these concerns, we aimed to explore the rising temperatures trends in NYC, WNV incidence in NYC, and what is being done to control its spread. 

  
## **Initial Questions**

With this inspiration, we wanted to explore available NYC data sets to answer some of these initial questions:

* What communities are affected most by rising temperatures?
* What do average temperatures look like over time in NYC? Is there any shift?
  * What about precipitation trends?
* If climate change is causing temperatures to rise in NYC, does this affect mosquito season?
* What is the incidence of mosquitoes carrying WNV in NYC over time? 
  * Is it different across boroughs?
* Is there a relationship between heat vulnerability index (HVI) score and incidence of WNV?
* Are the number of WNV human cases increasing over time in NYC?
  * If so, are there specific boroughs in NYC that are disproportionately affected by WNV?
* How has the NYC Department of Health and Mental Hygiene (DOHMH) mosquito control efforts during the summer affected the incidence of WNV case detection?

We hypothesize that ***between the years of 2021 to 2024, the NYC boroughs with warmer average temperatures and greater amounts of precipitation will have a higher incidence of mosquitoes carrying West Nile virus***. We also predict that areas with higher HVI scores will also have more WNV-positive mosquitoes as these areas may be more likely to experience worse heat conditions and potentially create more suitable environments for infected mosquitoes.
  
## **Data**

### Sources

We utilized the [Supplmentary Dataset of NYC Zip Codes](https://p8105.com/data/zip_codes.html) that was provided on our P8105 class website, which includes information on *NYC zip codes, boroughs*, and *neighborhoods*.

We located the [NYC Heat Vulnerability Index (HVI)](https://a816-dohbesp.nyc.gov/IndicatorPublic/data-explorer/climate/?id=2411#display=summary) through the NYC DOHMH Environment and Health Data Portal. We also utilized [Zip Code Tabulated Area Level HVI](https://catalog.data.gov/dataset/heat-vulnerability-index-rankings) data, which is designed to approximate zip codes and works well with merging HVI scores with our other datasets. The HVI shows the risk of community-level health impacts due to extreme heat and includes factors such as surface temperatures, green spaces, access to home air conditioning, median income, and the percentage of residents who are low-income or non-Latinx Black (who are often excluded from heat resources). Neighborhoods with a **HVI score of 1** have the lowest risk while those with a **HVI score of 5** have the highest risk.

Data from the [West Nile Virus Mosquito Activity reports](https://www.nyc.gov/site/doh/health/health-topics/west-nile-virus-activity.page) was used to show the detection of positive mosquitoes for West Nile virus in various NYC neighborhoods. Further, we used geographic files for [Modified ZIP Code Tabulation Areas (MODZCTA)](https://github.com/nychealth/coronavirus-data/tree/master/Geography-resources) to create the NYC maps. All data is provided by the NYC DOHMH. More information on how this mosquito surveillance is conducted by the NYC DOHMH can be found [here](https://www.nyc.gov/assets/doh/downloads/pdf/wnv/2023/wnvplan2023.pdf).

We located the [West Nile Virus NYC Human Cases](https://www.nyc.gov/assets/doh/downloads/pdf/wnv/2024/wnvplan2024.pdf) through the *2024 NYC DOHMH Comprehensive Mosquito Surveillance and Control Plan*. The report provides the number of West Nile virus human cases in NYC between 1999-2023, separated by borough.

The [Global Surface Summary of the Day (GSOD)](https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.ncdc:C00516) contains global daily average weather observations from weather stations. A `GSODR` package exists where we can load in data from specific stations and select various weather measurements. Our indicators include: mean temperature, dew point temperature, sea level pressure, station pressure, visibility, wind speed, gust, precipitation amount, and snow depth. This data is provided by the National Oceanic and Atmospheric Administration. 


### Cleaning

**P8105 Zip Codes Dataset**

The supplementary zip codes dataset was imported using `read_html` and initially cleaned using `janitor::clean_names`. We filtered the dataset to only include zip codes within NYC, used a mutate step to change borough names to: *Brooklyn*, *Manhattan*, and *Staten Island*, and only selected relevant variables: *zip_code*, *borough*, and *neighborhood*.

```{r, message=FALSE, echo=FALSE}
zip_nbhd <- read_html("https://p8105.com/data/zip_codes.html") |>
  html_table() |>
  first() |>
  janitor::clean_names() |>
  filter(!(county == "New York" & zip_code %in% c(10463, 11201))) |>#duplicated zip codes, both are *not* in New York County
  mutate(borough = str_replace_all(county, c(
    "Kings" = "Brooklyn",
    "New York" = "Manhattan",
    "Richmond" = "Staten Island"
  ))) |>
  select(zip_code, borough, neighborhood)
```

**Positive Mosquitoes Detected with West Nile Virus in NYC Datasets (2021-2024)**

For the years 2021-2023, the datasets provided by the NYC DOHMH are simple html tables that were imported using `read_html` and cleaned using `janitor::clean_names`. Zip codes for each detection of WNV-positive mosquitoes on a specific day are listed and separated by commas. To clean this, we used multiple `separate_longer_delim` steps to separate each positive mosquito detection into a separate observation, while keeping the corresponding detection date. 

For the year 2024, a .csv file was imported using `read_csv` and reshaped using `pivot_longer` to consolidate all WNV-positive mosquito detection dates into a single `date` column.

For all datasets, we reformatted the `zip_code` column to integers and the `date` column to a date object (yyyy-mm-dd), corrected any zip code and date entry errors, and removed unnecessary columns and N/A values. As a result, for each year, each row now represents a single WNV-positive mosquito detected in a specific zip code/borough/neighborhood on a specific date. 

```{r, message=FALSE, echo=FALSE, warning=FALSE}
## 2021 - 2023 are simple HTML tables
mosquitoes_2024_url <- "https://www.nyc.gov/site/doh/health/health-topics/west-nile-virus-activity-2024.page"
mosquitoes_2023_url <- "https://www.nyc.gov/site/doh/health/health-topics/west-nile-virus-activity-2023.page"
mosquitoes_2022_url <- "https://www.nyc.gov/site/doh/health/health-topics/west-nile-virus-activity-2022.page"
mosquitoes_2021_url <- "https://www.nyc.gov/site/doh/health/health-topics/west-nile-virus-activity-2021.page"

### 2024
mosquitoes_2024_table <- read_csv("data/mosquitoes_2024.csv") |>
  pivot_longer(cols = starts_with("date"), values_to = "date") |>
  select(-name, -detection_type) |>  
  mutate(
    date = case_when(
      date == "8/5/5024" ~ "8/5/2024",
      .default = date # one observation, 11426, with year miswritten as 5024-08-05 instead of 2024-08-05
    ),
    date = lubridate::mdy(date)
  ) |> 
  na.omit() 
   
### 2023
mosquitoes_2023_table <- mosquitoes_2023_url |>
  read_html() |> html_table() |> first() |>
  janitor::clean_names() |>
  separate_longer_delim(zip_code, delim = ", ") |>
  separate_longer_delim(zip_code, delim = ",\n") |>
  separate_longer_delim(zip_code, delim = ",") |> #this data is disgusting
  select(-neighborhood, -download) |>
  mutate(
    zip_code = as.integer(zip_code),
    date = str_c(date, ", ", 2023),
    date = lubridate::mdy(date)
  )

### 2022
mosquitoes_2022_table <- mosquitoes_2022_url |>
  read_html() |> html_table() |> first() |>
  janitor::clean_names() |>
  separate_longer_delim(zip_code, delim = ", ") |>
  separate_longer_delim(zip_code, delim = ",\n") |>
  separate_longer_delim(zip_code, delim = ",") |> #this data is disgusting
  select(-download) |>
  mutate(
    zip_code = as.integer(zip_code),
    date = str_c(date, ", ", 2022),
    date = lubridate::mdy(date)
  )

### 2021
mosquitoes_2021_table <- mosquitoes_2021_url |>
  read_html() |> html_table() |> first() |>
  janitor::clean_names() |>
  separate_longer_delim(zip_code, delim = ", ") |>
  separate_longer_delim(zip_code, delim = ",\n") |>
  separate_longer_delim(zip_code, delim = ",") |> #this data is disgusting
  select(-download) |>
  mutate(
    zip_code = as.integer(zip_code),
    date = str_c(date, ", ", 2021),
    date = lubridate::mdy(date)
  ) |> 
  mutate(
    zip_code = case_when(
      zip_code == 1123 ~ 11236, # data entry error on the website, where 1123 is missing the last digit - per https://www.nyc.gov/assets/doh/downloads/pdf/wnv/2021/wnv-activity-bk-08192021.pdf it should be 11236
    .default = zip_code
    )
  )
```

**NYC Positive Cases of West Nile Virus Dataset (1999-2023)**

Data was downloaded as a .csv file, imported using `read_csv`, and initially cleaned using `janitor::clean_names`. The data was reshaped using `pivot_longer` such that our columns are now: `year`, `borough`, and `wnv_cases` (i.e. the number of West Nile virus cases counted within that borough).

```{r, message=FALSE, echo=FALSE}
wnv_cases <- read_csv("data/wnv_cases.csv") |>
  janitor::clean_names() |>
  pivot_longer(cols = c(
    bronx,
    brooklyn,
    manhattan,
    queens,
    staten_island
  ),
  names_to = "borough",
  values_to = "wnv_cases") |> 
  mutate(
    borough = case_match(
      borough, 
      "manhattan" ~ "Manhattan",
      "brooklyn" ~ "Brooklyn",
      "queens" ~ "Queens",
      "bronx" ~ "Bronx",
      "staten_island" ~ "Staten Island"
    )
  )
```


**NYC Heat Vulnerability Index Dataset**

This dataset was already in a tidy format after downloading (thank you NYC DOHMH Environment and Health!!). We simply imported the dataset using `read_csv` and applied `janitor::clean_names` for consistency. Our columns include: `zip_codes` and `hvi` scores.

```{r, message=FALSE, echo=FALSE}
heat_vuln <- read_csv("data/heat_vulnerability.csv") |>
  janitor::clean_names()

hvi_zcta <- read_csv("https://data.cityofnewyork.us/resource/4mhf-duep.csv")  |> 
  rename(zip_code = zcta20)
write_csv(hvi_zcta, "cleaned_data/hvi_zcta.csv")
```


**GSODR Dataset**

Through online research, we identified Bushwick, Brooklyn as the geographical center of NYC. We loaded the `GSODR` package and used the `nearest_stations` function to identify weather stations surrounding the longitude and latitude coordinates of Bushwick. We used a 20-mile radius to capture stations across all boroughs. We then filtered out stations not in NYC and those without 2021-2024 weather data. This left us with only 5 stations: **Wall St**, **LGA**, **Central Park**, **JFK**, and **The Battery**.

We pulled weather data between 2021-2024 from the 5 weather stations using `get_GSOD`, applied `janitor::cleannames`, and selected these variables: `station name`, `latitude` and `longitude`, date (`yearmoda`) mean temperature (`temp`), mean precipitation (`prcp`), mean dew point (`dewp`) (to get an idea of water content in the air), and `max` and `min` temperature. Finally, we applied a series of mutate steps to convert temperature measurements to Fahrenheit, precipitation measures to inches, and renamed weather stations to more intuitive names.

```{r, eval=FALSE, echo=FALSE}
station_ids <-
  nearest_stations(
    LAT = 40.6958,
    LON = -73.9171,
    distance = 20
  ) |>
  filter(STATE != "NJ" & !(NAME %in% c("KINGS POINT",
                                       "SANDY HOOK",
                                       "BERGEN POINT",
                                       "MITCHEL FIELD",
                                       "HEMPSTEAD MITCHELL FLD AFB",
                                       "AMBROSE LIGHT  NY",
                                       "AMBROSE / FT. TILDEN",
                                       "FREEPORT"))) |>
  filter(!(STNID %in% c("999999-14732",
                        "725033-94728",
                        "725060-94728",
                        "999999-94728",
                        "744976-99999",
                        "999999-14786",
                        "999999-94789",
                        "725026-99999",
                        "997439-99999"))) |>
  janitor::clean_names() |>
  pull(stnid)

# weather_2021_2024 <-
#    get_GSOD(
#      years = c(2021:2024),
#      station = station_ids
#    ) |>
#    janitor::clean_names() |>
#    select(name, latitude, longitude, yearmoda, temp, max, min, prcp, dewp) |>
#    mutate(
#      temp = (temp * 9/5) + 32,
#      max = (max * 9/5) + 32,
#      min = (min * 9/5) + 32,
#      prcp = (prcp * 0.0393701),
#      year = year(yearmoda),
#      name = case_match(
#        name,
#        "PORT AUTH DOWNTN MANHATTAN WALL ST HEL" ~ "Wall St",
#        "LA GUARDIA AIRPORT" ~ "LGA",
#        "CENTRAL PARK" ~ "Central Park",
#        "JOHN F KENNEDY INTERNATIONAL AIRPORT" ~ "JFK",
#        "THE BATTERY" ~ "The Battery"
#      )
#    )

# Uncomment this line after running the previous block once
weather_2021_2024 <- read_csv("data/weather_2021_2023.csv")

#write_csv(weather_2021_2024, "data/weather_2021_2023.csv")
```


  
### Merging

To combine our datasets, we utilized the shared `zip_code` variable as a key for merging. This allowed us to successfully combine WNV-positive mosquito datasets (2021–2024), HVI scores, and MODZCTA shape files. For the mosquito data, we ensured that zip codes without mosquito detection were assigned a value of zero.

Our relevant variables for our datasets include:

**WNV+ Mosquito Data**

* `zip_code`: The zip code within NYC in which the WNV+ mosquitoes were detected
* `borough`: The NYC borough (Brooklyn, Manhattan, Bronx, Staten Island, or Queens) corresponding to the zip code location. 
* `date`: The date in which the WNV+ mosquito was detected within a specific zip code.
* `count`: The number of WNV+ mosquitoes detected within a specific zip code.
* `neighborhood`: The name of the NYC neighborhood in which the WNV+ mosquito was detected.
* `hvi`: The HVI score (ranging from 1-5)

```{r, message=FALSE, echo=FALSE, results='hide'}
wnv_mosquito_detection <- bind_rows(
  mosquitoes_2024_table,
  mosquitoes_2023_table,
  mosquitoes_2022_table,
  mosquitoes_2021_table,
) |> 
  left_join(y = zip_nbhd, by = join_by(zip_code)) |>
  select(zip_code, borough.x, date, neighborhood) |> 
  rename(borough = borough.x) |> # one ZIP code, 10000 has NA for borough as it is not in ZIP list - Central Park based on a Google search
  mutate(
    borough = case_when(
      zip_code == 10000 ~ "Manhattan",
      .default = borough
    ),
    neighborhood = case_when(
      zip_code == 10000 ~ "Central Park",
      .default = neighborhood
    )
  )

inner_join(wnv_mosquito_detection, hvi_zcta, by = join_by(zip_code)) |> 
  group_by(hvi) |> 
  summarize(
    count = n()
  )
```


**GSODR**

* `name`: The name of the NYC weather station
* `latitude`: Latitude coordinate of the weather station
* `longtitude`: Longitude coordinate of the weather station
* `yearmoda`: year of the weather station data that was pulled
* `temp`: mean temperature (Fahrenheit)
* `max`: maximum temperature measured (Fahrenheit)
* `min`: minimum temperature measured (Fahrenheit)
* `prcp`: mean precipitation (inches)
* `dewp`: mean dew point (Fahrenheit)

Our datasets are now ready for exploration and analysis!
  

  
  
  
  
## **Data Exploration**

### Comparing Temperature Data Across Weather Stations

We first wanted to look at which weather station to use for our temperature analysis as we were curious of whether there were any variations between temperature data from these 5 stations. We plotted the temperature data between 2021 and 2024. 

```{r, message=FALSE, echo=FALSE}
station_ids <-
  nearest_stations(
    LAT = 40.6958,
    LON = -73.9171,
    distance = 20
  ) |>
  filter(STATE != "NJ" & !(NAME %in% c("KINGS POINT",
                                       "SANDY HOOK",
                                       "BERGEN POINT",
                                       "MITCHEL FIELD",
                                       "HEMPSTEAD MITCHELL FLD AFB",
                                       "AMBROSE LIGHT  NY",
                                       "AMBROSE / FT. TILDEN",
                                       "FREEPORT"))) |>
  filter(!(STNID %in% c("999999-14732",
                        "725033-94728",
                        "725060-94728",
                        "999999-94728",
                        "744976-99999",
                        "999999-14786",
                        "999999-94789",
                        "725026-99999",
                        "997439-99999"))) |>
  janitor::clean_names() |>
  pull(stnid)

# weather_2021_2024 <-
#    get_GSOD(
#      years = c(2021:2024),
#      station = station_ids
#    ) |>
#    janitor::clean_names() |>
#    select(name, latitude, longitude, yearmoda, temp, max, min, prcp, dewp) |>
#    mutate(
#      temp = (temp * 9/5) + 32,
#      max = (max * 9/5) + 32,
#      min = (min * 9/5) + 32,
#      prcp = (prcp * 0.0393701),
#      year = year(yearmoda),
#      name = case_match(
#        name,
#        "PORT AUTH DOWNTN MANHATTAN WALL ST HEL" ~ "Wall St",
#        "LA GUARDIA AIRPORT" ~ "LGA",
#        "CENTRAL PARK" ~ "Central Park",
#        "JOHN F KENNEDY INTERNATIONAL AIRPORT" ~ "JFK",
#        "THE BATTERY" ~ "The Battery"
#      )
#    )

# Uncomment this line after running the previous block once
weather_2021_2024 <- read_csv("data/weather_2021_2023.csv")

#write_csv(weather_2021_2024, "data/weather_2021_2023.csv")

weather_2021_2024 |>
  ggplot(aes(x = yearmoda, y = temp, color = name)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(se = FALSE, color = "white", alpha = 0.5) +
  theme(legend.position = "bottom") + 
  geom_errorbar(aes(ymin = min, ymax = max), alpha = 0.5)
```

We found that, despite being in different boroughs, there is not much variation in measured temperatures. Since they were all fairly similar, we ended up taking the average of the 5 weather stations. 


### Exploring Trends in WNV-Positive Mosquito Counts

We then wanted to focus on the mosquito data. Specifically, we wanted to look at the number of mosquito counts by year and borough to explore whether mosquito season starts earlier. Since we predicted that temperatures are rising in NYC, then we should expect that warmer climates may cause mosquito season to start sooner. To visualize this, we created four separate plots of counts of mosquitoes positive for WNV for each year from 2021 through 2024, separated by borough. 


```{r, message=FALSE, warning=FALSE, echo=FALSE}
mosquitoes_2021_table |>
  group_by(borough, date) |> 
  summarize(
    count = n()
  ) |> 
  ggplot(aes(x = date, y = count, fill = borough)) +
  geom_col(stat = "identity") +
  facet_grid(~borough)

mosquitoes_2022_table |>
  group_by(borough, date) |> 
  summarize(
    count = n()
  ) |> 
  ggplot(aes(x = date, y = count, fill = borough)) +
  geom_col(stat = "identity") +
  facet_grid(~borough)

mosquitoes_2023_table |>
  group_by(borough, date) |> 
  summarize(
    count = n()
  ) |> 
  ggplot(aes(x = date, y = count, fill = borough)) +
  geom_col(stat = "identity") +
  facet_grid(~borough)

mosquitoes_2024_table |>
  group_by(borough, date) |> 
  summarize(
    count = n()
  ) |> 
  ggplot(aes(x = date, y = count, fill = borough)) +
  geom_col(stat = "identity") +
  facet_grid(~borough)
```

The figures show that there are drastic differences in the counts of WNV-positive mosquitoes between boroughs. Queens consistently had the highest counts of WNV-positive mosquitoes compared to the other boroughs across all years, while Manhattan had the lowest counts of WNV-positive mosquitoes. 

```{r, message=FALSE, warning=FALSE, echo=FALSE}
wnv_mosquito_detection |> 
  mutate(count = 1) |> 
  group_by(date) |> 
  summarize(
    count = sum(count)
  ) |> 
  ggplot(aes(x = date, y = count)) +
  geom_point()
```

Further, when boroughs are grouped together, we see that the number of WNV-positive mosquitoes peaked in 2021, followed by a decline in 2022 and 2023. There is a slight increase in the counts peak in the year 2024.

We can also see from both figures that between the years of 2021-2024, mosquito season typically ranges from June to October, which is still within a normal range of months for mosquito season in NYC. Unfortunately, the data we have access to only provides detections of WNV+ mosquitoes within NYC starting from 2021 so we won't be able to analyze long-term trends.


### Mapping WNV-Positive Mosquito Trends Across NYC

To visualize this further, we wanted to create a map of NYC between 2021-2024 and overlay the WNV positive mosquitoes data to see if we could visualize any trends on which boroughs or neighborhoods may be affected more. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# get ZIP codes that did not have counts of WNV using anti_join with original ZIP code list (for generating proportions and to ensure ZIPs without cases can still be mapped later)


all_zctas <-
  zip_nbhd |> 
  pull(zip_code) |> 
  c(10000) |> 
  expand_grid(
    zip_code = _,
    year = c(2021, 2022, 2023, 2024),
    count = 0
  )

fill_neg_zctas <- function(yr) {
  
  pos = 
    wnv_mosquito_detection |> 
    mutate(
      year = year(date)
    ) |> 
    filter(year == yr) |> 
    group_by(zip_code, neighborhood, year) |> 
    summarize(
      count = n()
    ) |>
    arrange(count)
  
  neg = 
    all_zctas |> 
    filter(year == yr) |> 
    anti_join(pos, by = join_by(zip_code))
  
  rbind(pos, neg)
    
}

wnv_mosquito_data <-
  tibble(
    years = c(2021, 2022, 2023, 2024),
    mosquito_data = map(years, fill_neg_zctas)
  ) |> 
  unnest(mosquito_data) |> 
  select(-years) |> 
  left_join(zip_nbhd, by = join_by(zip_code, neighborhood)) |> # rejoin borough variable from original ZIP code list 
  mutate(
    borough = case_when(
      zip_code == 10000 ~ "Manhattan",
      .default = borough
    ),
    neighborhood = case_when(
      zip_code == 10000 ~ "Central Park",
      .default = neighborhood
    )
  )

wnv_mosquito_data |> 
  pull(count) |> 
  sum() # 2924

wnv_mosquito_detection |> 
  nrow() # 2924 - matches

# get geographic coordinates for use in ggplot later

modzcta_path <- "data/shapefiles/MODZCTA_2010.shp"

zip_shp <-
  modzcta_path |>
  st_read() |> 
  drop_na(label) |> # gets rid of MODZCTA 99999 which is not a real MODZCTA 
  separate_longer_delim(label, delim = ", ") |> 
  janitor::clean_names() |> 
  mutate(
    modzcta = as.double(modzcta),
    zip_code = as.double(label)
  ) |> 
  separate_longer_delim(label, delim = ", ")

# get MODZCTA for each ZIP code (approximating ZIP code as ZCTA)

mos_by_modzcta <- 
  left_join(wnv_mosquito_data, zip_shp, by = join_by("zip_code")) |> 
  mutate(
    neighborhood = case_when(
      is.na(neighborhood) ~ "Non-residential",
      .default = neighborhood
    )
  )

mos_by_modzcta |>
  ggplot(aes(fill = count, geometry = geometry)) +
  geom_sf() + 
  labs(
    title = "WNV positive mosquitoes by MODZCTA",
    caption = "Data and shapefile from DOHMH"
  ) +
  theme(
    axis.text.x = element_blank(), 
    axis.title.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank()
  ) +
  viridis::scale_fill_viridis(limits = c(0, 25)) +
  facet_wrap(~year)
```

Interestingly, from the figure, we found that there are zip codes in Queens and Staten Island that appear to have an increase in the number of mosquitoes positive for WNV from 2021 to 2024. Manhattan, Brooklyn, and the Bronx remained relatively low.  





  
## **Data Analysis**
formal statistical analyses  
can link to page with full code  
  
## **Discussion**
*findings  
*are they what we expected?  
*what insights into data can we make?   


